---
title: 'NFL Referee Penalties: Data Loading and EDA'
author: "Jiashen Wang"
date: "`r Sys.Date()`"
output: pdf_document
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4,
  dpi = 140
)
```

- use `nflreadr::load_pbp()` for penalty events and game context,
- use `nflreadr::load_officials()` for game-level official assignments,

```{r libraries, include=FALSE}
library(tidyverse)
library(nflreadr)
library(janitor)
library(skimr)
library(gt)
library(scales)
library(lubridate)
library(glue)
library(forcats)
library(patchwork)
library(broom)
```

```{r parameters}
seasons <- 2015:2025
season_type_keep <- c("REG")
```


# 1) Load datasets

```{r load-pbp}
pbp_raw <- load_pbp(seasons) %>%
  clean_names()

pbp_raw %>% dim()
```

```{r load-officials}
officials_raw <- load_officials(seasons) %>%
  clean_names()

officials_raw %>% dim()
```

```{r officials-glimpse}
head(officials_raw)
```

# 2) EDA: Play-by-play data

```{r pbp-clean}
pbp <- pbp_raw %>%
  filter(season %in% seasons) %>%
  filter(if ("season_type" %in% names(.)) season_type %in% season_type_keep else TRUE) %>%
  filter(!is.na(game_id))

pbp %>% summarise(rows = n(), games = n_distinct(game_id), seasons = n_distinct(season))
```

```{r pbp-head}
pbp %>%
  select(c("game_id", "season", "week", "home_team", "away_team", "posteam", "defteam",
                "qtr", "down", "play_type", "desc", "penalty", "penalty_yards")) %>%
  slice_head(n = 5)
```

## Identify penalty plays

```{r penalty-definition}
penalty_plays <- pbp %>%
  filter(if ("penalty" %in% names(.)) penalty == 1 else FALSE)

penalty_plays %>%
  summarise(
    penalty_rows = n(),
    penalty_games = n_distinct(game_id),
    pct_of_all_rows = n() / nrow(pbp)
  )
```

### Penalty rate by season

```{r penalty-by-season-table}
penalty_by_season <- pbp %>%
  group_by(season) %>%
  summarise(
    plays = n(),
    penalties = sum(penalty == 1, na.rm = TRUE),
    penalty_rate = penalties / plays,
    avg_penalty_yards = mean(penalty_yards[penalty == 1], na.rm = TRUE),
    .groups = "drop"
  )

penalty_by_season %>%
  mutate(
    penalty_rate = percent(penalty_rate, accuracy = 0.01),
    avg_penalty_yards = round(avg_penalty_yards, 2)
  ) %>%
  gt()
```

```{r penalty-by-season-plot}
ggplot(penalty_by_season, aes(x = season, y = penalty_rate)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  ylim(0.05, 0.08) +
  labs(
    title = "Penalty play rate by season",
    x = "Season",
    y = "Penalty plays / all play-by-play rows"
  ) +
  theme_minimal()
```

### Distribution of penalty yards

```{r penalty-yards-hist}
penalty_plays %>%
  filter(!is.na(penalty_yards), penalty_yards >= 0) %>%
  ggplot(aes(x = penalty_yards)) +
  geom_histogram(binwidth = 5) +
  labs(
    title = "Distribution of penalty yards on flagged plays",
    x = "Penalty yards",
    y = "Count"
  ) +
  theme_minimal()
```

### Which situations produce more penalties?

```{r penalty-situation-summary}
penalty_context <- pbp %>%
  mutate(
    is_penalty = penalty == 1,
    qtr = factor(qtr),
    down = factor(down),
    play_type = fct_lump_n(as.factor(play_type), n = 8)
  )

context_summary <- penalty_context %>%
  group_by(qtr, down, play_type) %>%
  summarise(
    plays = n(),
    penalties = sum(is_penalty, na.rm = TRUE),
    penalty_rate = penalties / plays,
    .groups = "drop"
  ) %>%
  filter(plays >= 100)

context_summary %>%
  arrange(desc(penalty_rate)) %>%
  slice_head(n = 10) %>%
  gt()
```

```{r penalty-by-quarter}
pbp %>%
  mutate(qtr = factor(qtr)) %>%
  group_by(qtr) %>%
  summarise(
    plays = n(),
    penalties = sum(penalty == 1, na.rm = TRUE),
    rate = penalties / plays,
    .groups = "drop"
  ) %>%
  ggplot(aes(x = qtr, y = rate)) +
  geom_col() +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Penalty rate by quarter",
    x = "Quarter",
    y = "Penalty rate"
  ) +
  theme_minimal()
```

```{r penalty-by-down}
pbp %>%
  filter(!is.na(down)) %>%
  mutate(down = factor(down)) %>%
  group_by(down) %>%
  summarise(
    plays = n(),
    penalties = sum(penalty == 1, na.rm = TRUE),
    rate = penalties / plays,
    .groups = "drop"
  ) %>%
  ggplot(aes(x = down, y = rate)) +
  geom_col() +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Penalty rate by down",
    x = "Down",
    y = "Penalty rate"
  ) +
  theme_minimal()
```

### Most common penalty teams and penalty types

```{r penalty-team-type-tables}
penalty_plays %>%
  count(penalty_team, sort = TRUE) %>%
  slice_head(n = 15) %>%
  gt()
```

```{r desc-sample-penalties}
penalty_plays %>%
  select(c("game_id", "season", "week", "posteam", "defteam", "play_type", "desc", "penalty_yards")) %>%
  slice_head(n = 5)
```

## Game-level aggregation from play-by-play

```{r game-level-pbp}
game_level <- pbp %>%
  group_by(game_id, old_game_id, season, week, season_type, home_team, away_team) %>%
  summarise(
    total_rows = n(),
    total_penalties = sum(penalty == 1, na.rm = TRUE),
    total_penalty_yards = sum(penalty_yards, na.rm = TRUE),
    unique_posteams = n_distinct(posteam, na.rm = TRUE),
    mean_abs_score_diff = mean(abs(score_differential), na.rm = TRUE),
    mean_ydstogo = mean(ydstogo, na.rm = TRUE),
    pass_rate = mean(play_type == "pass", na.rm = TRUE),
    rush_rate = mean(play_type == "run", na.rm = TRUE),
    .groups = "drop"
  )
```

```{r game-penalty-distribution}
ggplot(game_level, aes(x = total_penalties)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of total penalties per game",
    x = "Total penalties in game",
    y = "Number of games"
  ) +
  theme_minimal()
```

```{r game-overdispersion-check}
game_level %>%
  summarise(
    mean_penalties = mean(total_penalties, na.rm = TRUE),
    var_penalties = var(total_penalties, na.rm = TRUE),
    overdispersion_ratio = var_penalties / mean_penalties
  ) %>%
  gt()
```

## Team-game aggregation from play-by-play

```{r team-game-level}
team_game_level <- pbp %>%
  filter(!is.na(posteam)) %>%
  group_by(game_id, old_game_id, season, week, season_type, home_team, away_team, posteam) %>%
  summarise(
    team_plays = n(),
    team_penalties = sum(penalty == 1 & penalty_team == first(posteam), na.rm = TRUE),
    team_penalty_yards = sum(if_else(penalty_team == first(posteam), penalty_yards, 0), na.rm = TRUE),
    offensive_penalty_rows = sum(penalty == 1 & posteam == penalty_team, na.rm = TRUE),
    mean_ydstogo = mean(ydstogo, na.rm = TRUE),
    mean_abs_score_diff = mean(abs(score_differential), na.rm = TRUE),
    early_down_rate = mean(down %in% c(1, 2), na.rm = TRUE),
    pass_rate = mean(play_type == "pass", na.rm = TRUE),
    rush_rate = mean(play_type == "run", na.rm = TRUE),
    is_home = as.integer(first(posteam) == first(home_team)),
    .groups = "drop"
  )
```

```{r home-away-penalties}
team_game_level %>%
  mutate(home_away = if_else(is_home == 1, "Home", "Away")) %>%
  group_by(home_away) %>%
  summarise(
    mean_penalties = mean(team_penalties, na.rm = TRUE),
    median_penalties = median(team_penalties, na.rm = TRUE),
    mean_penalty_yards = mean(team_penalty_yards, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  gt()
```

```{r home-away-boxplot}
team_game_level %>%
  mutate(home_away = if_else(is_home == 1, "Home", "Away")) %>%
  ggplot(aes(x = home_away, y = team_penalties)) +
  geom_boxplot() +
  labs(
    title = "Distribution of team penalties by home/away status",
    x = NULL,
    y = "Team penalties in game"
  ) +
  theme_minimal()
```


# 3) EDA: Officials data

```{r officials-basic-summary}
officials <- officials_raw %>%
  filter(season %in% seasons) %>%
  filter(if ("season_type" %in% names(.)) season_type %in% season_type_keep else TRUE)

officials %>%
  summarise(
    rows = n(),
    games = n_distinct(game_id),
    officials = n_distinct(official_name),
    positions = n_distinct(position)
  )
```

```{r officials-head}
officials %>%
  slice_head(n = 5)
```

## Extract lead referee summaries

```{r referee-game-level}
referee_game <- officials %>%
  filter(position == "Referee") %>%
  transmute(
    game_id,
    season,
    week,
    lead_referee = official_name,
    lead_referee_id = official_id,
    referee_jersey = jersey_number
  )

referee_game %>%
  summarise(games = n(), unique_referees = n_distinct(lead_referee))
```

```{r top-referees-by-games}
referee_game %>%
  count(lead_referee, sort = TRUE) %>%
  slice_head(n = 10) %>%
  gt()
```


# 4) Joined EDA: penalties merged with officiating data

```{r}
referee_game <- officials %>%
  filter(position == "Referee") %>%
  transmute(
    old_game_id = game_id,
    season,
    week,
    lead_referee = official_name,
    lead_ref_id = official_id
  )
```

```{r joined-game-level}
game_joined <- game_level %>%
  left_join(referee_game, by = c("old_game_id", "season", "week"))

game_joined %>%
  summarise(
    games = n(),
    missing_referee = sum(is.na(lead_referee)),
    pct_missing_referee = mean(is.na(lead_referee))
  )
```

```{r joined-game-head}
game_joined %>%
  select(c("game_id", "season", "week", "home_team", "away_team",
                "total_penalties", "total_penalty_yards", "total_rows",
                "lead_referee")) %>%
  slice_head(n = 5)
```

## Team-game joined dataset

```{r joined-team-game}
team_game_joined <- team_game_level %>%
  left_join(referee_game, by = c("old_game_id", "season", "week"))

table(
  has_referee = !is.na(team_game_joined$lead_referee),
  useNA = "ifany"
)
```

## Raw referee differences: do some refs call more penalties?

### Referee-level game averages

```{r ref-summary-game}
ref_game_summary <- game_joined %>%
  filter(!is.na(lead_referee)) %>%
  group_by(lead_referee) %>%
  summarise(
    games = n(),
    mean_penalties = mean(total_penalties, na.rm = TRUE),
    median_penalties = median(total_penalties, na.rm = TRUE),
    sd_penalties = sd(total_penalties, na.rm = TRUE),
    mean_penalty_yards = mean(total_penalty_yards, na.rm = TRUE),
    mean_total_rows = mean(total_rows, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(games >= 8) %>%
  arrange(desc(mean_penalties))

ref_game_summary %>%
  slice_head(n = 20) %>%
  gt()
```

```{r ref-avg-penalties-plot}
ref_game_summary %>%
  slice_max(order_by = mean_penalties, n = 20) %>%
  mutate(lead_referee = fct_reorder(lead_referee, mean_penalties)) %>%
  ggplot(aes(x = lead_referee, y = mean_penalties)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Highest raw average penalties per game by lead referee",
    x = NULL,
    y = "Mean penalties per game"
  ) +
  theme_minimal()
```

### Relationship between game pace and penalties, colored by referee

```{r plays-vs-penalties, fig.width=8, fig.height=6}
set.seed(123)
plot_data <- game_joined %>%
  filter(!is.na(lead_referee)) %>%
  mutate(lead_referee = fct_lump_n(lead_referee, n = 10))

ggplot(plot_data, aes(x = total_rows, y = total_penalties, color = lead_referee)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Games with more plays tend to have more penalties",
    x = "Total play-by-play rows in game",
    y = "Total penalties"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


```{r}
ggplot(game_joined, aes(x = total_rows, y = total_penalties)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "Relationship between total plays and penalties in NFL games",
    x = "Number of plays in a game",
    y = "Number of penalties"
  ) +
  theme_minimal()
```


### Home / away penalty differences under different referees

```{r referee-home-away}
ref_home_away <- team_game_joined %>%
  filter(!is.na(lead_referee)) %>%
  mutate(home_away = if_else(is_home == 1, "Home", "Away")) %>%
  group_by(lead_referee, home_away) %>%
  summarise(
    games = n(),
    mean_team_penalties = mean(team_penalties, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(games >= 8)

ref_home_away %>%
  slice_head(n = 10) %>%
  gt()
```

## Team and referee interaction patterns

```{r team-ref-heatmap-data}
team_ref_summary <- team_game_joined %>%
  filter(!is.na(lead_referee)) %>%
  group_by(lead_referee, posteam) %>%
  summarise(
    games = n(),
    mean_penalties = mean(team_penalties, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(games >= 3)

team_ref_summary %>%
  arrange(desc(games)) %>%
  slice_head(n = 10) %>%
  gt()
```

```{r ref-team-heatmap}
heatmap_data <- team_ref_summary %>%
  group_by(lead_referee) %>%
  summarise(total_games = sum(games), .groups = "drop") %>%
  slice_max(order_by = total_games, n = 12) %>%
  select(lead_referee) %>%
  inner_join(team_ref_summary, by = "lead_referee")

ggplot(heatmap_data, aes(x = posteam, y = fct_reorder(lead_referee, games, .fun = sum), fill = mean_penalties)) +
  geom_tile() +
  labs(
    title = "Mean team penalties by team and lead referee",
    x = "Team",
    y = "Lead referee",
    fill = "Mean penalties"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
